---
title: "02-practical-code"
author: "Kaitlyn Johnson"
date: "2025-07-08"
output: html_document
---
# Load packages
```{r}
library(epiparameter)
library(EpiNow2)
library(tidyverse)
```

# Transmission: Activity 1
Estimate R(t), new infections, and doubling/halfing time using:
- Incidence of reported cases per day
- Reporting delay

1. What phase of the epidemic are you observing? (Exponential growth phase, near peak, or decay end phase)
2. Is the expected change in daily reports consistent with the estimated effective reproductive number, growth rate, and doubling time?
3. Interpret: How would you communicate these results to a decision-maker?
4. Compare: What differences do you identify from other group outputs? (if available)


# Read reported cases for ebola -----------------------------------------------
Ebola 35 days
```{r}
dat <- read_rds(
  "https://epiverse-trace.github.io/tutorials-middle/data/ebola_35days.rds"
) %>%
  dplyr::select(date, confirm = cases)
```

# Define a generation time from {epiparameter} to {EpiNow2} ---------------
Here we are using the serial interval as an approximation to the GI (remember 
serial interval is time between symptom onset of infector to symptom onset of 
infectee, generation interval is time between incident infection of 
infector to infectee.) The GI is always positive, whereas the serial interval
can be negative. 
```{r}
# access a serial interval
dat_serialint <- epiparameter::epiparameter_db(
  disease = "ebola",
  epi_name = "serial",
  single_epiparameter = TRUE
)
```
This is an object(list) containing disease, pathogen, epi parameters, study
citation, distribution and parameters of that distribution. 

# Extract the parameters from the epiparameter object
```{r}
# extract parameters from {epiparameter} object
dat_serialint_params <- epiparameter::get_parameters(dat_serialint)
```
This is a named vector.

Pass the parameters from epiparameter into the distribution interface in 
`EpiNow2`. Can either use the shape and scale parameters or the mean and
sd. Alternatively, if you knew only the mean and sd or shape and scale you could
just pass them in directly
```{r}
# adapt {epiparameter} to {EpiNow2} distribution inferfase
dat_generationtime <- EpiNow2::Gamma(
  shape = dat_serialint_params["shape"],
  scale = dat_serialint_params["scale"]
)
# or
dat_generationtime <- EpiNow2::Gamma(
  mean = dat_serialint$summary_stats$mean,
  sd = dat_serialint$summary_stats$sd
)
```
# Define the delays from infection to case report for {EpiNow2} -----------
This will be the delay from infection to sympton onset (also know as the
incubation period) and the time from symptom onset to case report. EpiNow2 
handles combining these two delay distributions under the hood (with a 
convolution). 

# Start by defining manually a delay from symptom onset to case report using 
the specified parameters.
```{r}
# define delay from symptom onset to case report
dat_reportdelay <- EpiNow2::LogNormal(
  meanlog = EpiNow2::Normal(mean = 1.4, sd = 0.5),
  sdlog = EpiNow2::Normal(mean = 0.25, sd = 0.2),
  max = 5
)
```

Use epiparameter to get the incubation time. 
```{r}
# define a delay from infection to symptom onset
dat_incubationtime <- epiparameter::epiparameter_db(
  disease = "ebola",
  epi_name = "incubation",
  single_epiparameter = TRUE
)

# incubation period: extract distribution parameters
dat_incubationtime_params <- epiparameter::get_parameters(
  dat_incubationtime
)
```

Get the maximum value of the incubation period because we are going to need to 
set a maximum in EpiNow2
```{r}
# incubation period: discretize and extract maximum value (p = 99%)
dat_incubationtime_max <- dat_incubationtime %>% 
  epiparameter::discretise() %>%
  quantile(p = 0.99)
# or
dat_incubationtime_max <- dat_incubationtime %>%
  quantile(p = 0.99) %>%
  base::round()

# incubation period: adapt to {EpiNow2} distribution interface
dat_incubationtime_epinow <- EpiNow2::Gamma(
  shape = dat_incubationtime_params["shape"],
  scale = dat_incubationtime_params["scale"],
  max = dat_incubationtime_max
)
```

# print required input
```{r}
dat_generationtime
dat_reportdelay
dat_incubationtime_epinow
```

# Set the number of parallel cores for {EpiNow2} --------------------------
```{r}
withr::local_options(list(mc.cores = parallel::detectCores() - 1))
```

# Estimate transmission using EpiNow2::epinow() ---------------------------
# with EpiNow2::*_opts() functions for generation time, delays, and stan.
```{r}
estimates <- EpiNow2::epinow(
  data = dat,
  generation_time = EpiNow2::generation_time_opts(dat_generationtime),
  delays = EpiNow2::delay_opts(dat_incubationtime_epinow + dat_reportdelay),
  stan = EpiNow2::stan_opts(samples = 1000, chains = 3)
)
```

# Print plot and summary table outputs ------------------------------------
```{r}
summary(estimates)
plot(estimates)

```
# Repeat for COVID 30 days
```{r}
dat <- read_rds(
  "https://epiverse-trace.github.io/tutorials-middle/data/covid_30days.rds"
) %>%
  dplyr::select(date, confirm)

```
# Define a generation time from {epiparameter} to {EpiNow2} ---------------

# access a serial interval (again using serial interval as a GI approx)
```{r}
dat_serialint <- epiparameter::epiparameter_db(
  disease = "covid",
  epi_name = "serial",
  single_epiparameter = TRUE
)
```

# extract parameters from {epiparameter} object
```{r}
dat_serialint_params <- epiparameter::get_parameters(dat_serialint)

# adapt {epiparameter} to {EpiNow2} distribution inferfase
dat_generationtime <- EpiNow2::LogNormal(
  meanlog = dat_serialint_params["meanlog"],
  sdlog = dat_serialint_params["sdlog"]
)
# or
dat_generationtime <- EpiNow2::LogNormal(
  mean = dat_serialint$summary_stats$mean,
  sd = dat_serialint$summary_stats$sd
)
```

# Define the delays from infection to case report for {EpiNow2} -----------
reporting delay (we were told this)
```{r}
# define delay from symptom onset to case report
dat_reportdelay <- EpiNow2::Gamma(
  mean = EpiNow2::Normal(mean = 2, sd = 0.5),
  sd = EpiNow2::Normal(mean = 1, sd = 0.5),
  max = 5
)
```
incubation time 
```{r}
# define a delay from infection to symptom onset
dat_incubationtime <- epiparameter::epiparameter_db(
  disease = "covid",
  epi_name = "incubation",
  single_epiparameter = TRUE
)

# incubation period: extract distribution parameters
dat_incubationtime_params <- epiparameter::get_parameters(
  dat_incubationtime
)

# incubation period: discretize and extract maximum value (p = 99%)
dat_incubationtime_max <- dat_incubationtime %>%
  epiparameter::discretise() %>%
  quantile(p = 0.99)
# or
dat_incubationtime_max <- dat_incubationtime %>%
  quantile(p = 0.99) %>%
  base::round()

# incubation period: adapt to {EpiNow2} distribution interface
dat_incubationtime_epinow <- EpiNow2::LogNormal(
  meanlog = dat_incubationtime_params["meanlog"],
  sdlog = dat_incubationtime_params["sdlog"],
  max = dat_incubationtime_max
)

# print required input
dat_generationtime
dat_reportdelay
dat_incubationtime_epinow
```

# Set the number of parallel cores for {EpiNow2} --------------------------
```{r}
withr::local_options(list(mc.cores = parallel::detectCores() - 1))


# Estimate transmission using EpiNow2::epinow() ---------------------------
# with EpiNow2::*_opts() functions for generation time, delays, and stan.
estimates <- EpiNow2::epinow(
  data = dat,
  generation_time = EpiNow2::generation_time_opts(dat_generationtime),
  delays = EpiNow2::delay_opts(dat_reportdelay + dat_incubationtime_epinow),
  stan = EpiNow2::stan_opts(samples = 1000, chains = 3)
)
```

# Print plot and summary table outputs ------------------------------------
```{r}
summary(estimates)
plot(estimates)
```

For the other days, simply swap out the path to the data for covid and ebola for 60 days and then interpret.


# Severity: Activity 2
Estimate the naive CFR (nCFR) and delay-adjusted CFR (aCFR) using the following inputs:
- Reported cases (aggregate incidence by date of onset)
- Onset to death delay

Questions:

1. What phase of the epidemic are you observing? (Exponential growth phase, near peak, or decay end phase)
2. Does the time series include all the possible deaths to observe from known cases?
3. How much difference is there between the nCFR and aCFR estimates?
4. Interpret: How would you communicate these results to a decision-maker?
5. Compare: What differences do you identify from other group outputs? (if available)

# Load packages -----------------------------------------------------------

```{r}
library(cfr)
library(epiparameter)
library(tidyverse)
```

# Read reported cases -----------------------------------------------------
```{r}
disease_dat <- readr::read_rds(
  "https://epiverse-trace.github.io/tutorials-middle/data/diamond_70days.rds"
)

disease_dat
```
Create incidence object (this pivots the dataframe from wide to long). It will
replace any missing values of dates with 0s.
```{r}
disease_incidence <- disease_dat %>%
  incidence2::incidence(
    date_index = "date",
    counts = c("cases", "deaths"),
    complete_dates = TRUE
  )

plot(disease_incidence)
```
# Confirm {cfr} data input format ----------------------------------------

# Is the input data already adapted to {cfr} input? This wants 
```{r}
disease_adapted <- disease_dat
# # OR
# # Does the input data need to be adapted to {cfr}? 
# disease_adapted <- disease_incidence %>%
#   cfr::prepare_data(
#     cases_variable = "cases",
#     deaths_variable = "deaths",
#     fill_NA = TRUE
#   )

disease_adapted

```

Get delay distributions from epi parameter (same as before) 
```{r}
disease_delay <- epiparameter::epiparameter_db(
  disease = "covid",
  epi_name = "onset_to_death",
  single_epiparameter = TRUE
)

```

# Estimate naive and adjusted CFR ----------------------------------------
```{r}
# Estimate static CFR
disease_adapted %>%
  filter(
    date < ymd(20200301)
  ) %>%
  cfr::cfr_static()
```
Compare to naive calculation
```{r}
disease_adapted |>
  filter(date < ymd("2020-03-01")) |>
  summarise(cfr = sum(deaths)/sum(cases)) |> 
  pull(cfr)
```


```{r}
disease_adapted %>%
  filter(
    date < ymd(20200301)
  ) %>%
  cfr::cfr_static(
    delay_density = function(x) density(disease_delay, x)
  )
```

# Estimate using complete time series ---------------------------------------
```{r}
# Estimate static CFR
disease_adapted %>%
  cfr::cfr_static()

# Estimate static delay-adjusted CFR
disease_adapted %>%
  cfr::cfr_static(
    delay_density = function(x) density(disease_delay, x)
  )
```

Repeat for MERS
```{r}
disease_dat <- readr::read_rds(
  "https://epiverse-trace.github.io/tutorials-middle/data/mers_linelist.rds"
)

disease_dat
```
# Create incidence object ------------------------------------------------
```{r}

disease_incidence <- disease_dat %>%
  incidence2::incidence(
    date_index = c("dt_onset", "dt_death"),
    complete_dates = TRUE
  )

plot(disease_incidence)
```

# Confirm {cfr} data input format ----------------------------------------
```{r}
# Is the input data already adapted to {cfr} input?  NO
disease_adapted <- disease_dat
# OR This needs to be run
# Does the input data need to be adapted to {cfr}? 
disease_adapted <- disease_incidence %>%
  cfr::prepare_data(
    cases = "dt_onset",
    deaths = "dt_death"
  )

disease_adapted
```

# Access delay distribution from epiparameter--------------------------------
```{r}
disease_delay <- epiparameter::epiparameter_db(
  disease = "mers",
  epi_name = "onset-to-death",
  single_epiparameter = TRUE
)
```
# Estimate naive and adjusted CFR ----------------------------------------

# Estimate static CFR
```{r}
disease_adapted %>%
  cfr::cfr_static()

```
# Estimate static delay-adjusted CFR
```{r}
disease_adapted %>%
  cfr::cfr_static(
    delay_density = function(x) density(disease_delay, x)
  )
```
