---
title: "03-practical-code.Rmd"
author: "Kaitlyn Johnson"
date: "2026-07-15"
output: html_document
---

Week 3: Estimate superspreading and simulate transmission chains

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(EpiNow2)
library(epiparameter)
library(cfr)
library(outbreaks)
library(epicontacts)
library(fitdistrplus)
library(superspreading)
library(epichains)
```

# Activity 1: Account for superspreading

Goal is to estimate the extent of individual-level variation (i.e. the dispersion parameter) of the offspring distribution -- the number of secondary cases per individual, and assess the implications for decision-making using line list cases and contact tracing data.

Questions: 
- Does the estimated dispersion parameter correlate with the contact
  network and histogram of secondary cases?
- What is the probability of new cases originating from a cluster of at
  least 10 cases?
- Would you recommend a backward contact tracing strategy?
- Does your dispersion parameter suggest a high or low risk of
  superspreading? Is this epidemic more likely to explode or be
  controlled?

Read linelis and contact data
```{r}
room_number <- 1

dat_contacts <- readr::read_rds(
  "https://epiverse-trace.github.io/tutorials-middle/data/set-01-contacts.rds"  #<DIFFERENT PER ROOM>
)

dat_linelist <- readr::read_rds(
  "https://epiverse-trace.github.io/tutorials-middle/data/set-01-linelist.rds"  #<DIFFERENT PER ROOM>
)

```

Create an epicontacts object
This is a directed epicontacts network 
```{r}
epi_contacts <- epicontacts::make_epicontacts(
  linelist = dat_linelist,
  contacts = dat_contacts,
  directed = TRUE
)
epi_contacts
```

Visualise the contact network 
```{r}
contact_network <- epicontacts::vis_epicontacts(epi_contacts)
contact_network
```
 
 Count the number of secondary cases per subject in contacts and linelist.
 First calculate the *out-degree* per case from all linelist cases
 
```{r}
 secondary_cases <- epicontacts::get_degree(
  x = epi_contacts,
  type = "out",
  only_linelist = TRUE
)
```

Plot the histogram of secondary cases
```{r}
individual_reproduction_num <- secondary_cases %>%
  enframe() %>% 
  ggplot(aes(value)) +
  geom_histogram(binwidth = 1) +
  labs(
    x = "Number of secondary cases",
    y = "Frequency"
  )

# Print output
individual_reproduction_num
```